Stock Market Valuation by Shawn P Emhe II
========================================================


```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load packages

library(ggplot2)
library(gridExtra)
library(ggthemes)
library(gsubfn)
library(rpart)
library(memisc)
```

In *Irrational Exhuberance*, Robert Shiller compiled over 100 years of data to
demonstrate how market dynamics create cycles in which the stock market prices
become disconnected with valuations. One criticism[^1] of his work is that
prices are set by the market, not by valuations, and because of this no one can
say when the market is "overvalued" or "undervalued." I intend to use the 
dataset to explore the link between value and price.

[^1]: [StockCharts Irrational Exuberance Review](
http://stockcharts.com/school/doku.php?id=chart_school:overview:
irrational_exuberanc)

```{r echo=FALSE, Load_the_Data}
# Load the Data
cape <- read.csv("ie_data.csv", sep = "\t")
```

# Data Wrangling

The data is already in tidy format, but still requires some wrangling prior to
exploration.

```{r echo=FALSE, Wrangle_Data}
str(cape)
```

The `str` command shows that there are two date fields.

1. Date: `r cape$Date[1:3]` ...
2. DateFraction: `r cape$DateFraction[1:3]` ...

Both use a YYYY.MM format, with the first numbering the months 01 through 12 and
the second representing months as a fraction of the year. I reviewed the 
background information[^2] on the dataset and the original excel file. The most
logical explanation I could find was that the second format was created for use
as the X-axis for the charts in the excel file. The first format was likely an
output of another program or data source and is not a format interpretable by
excel.

I converted the values into an R friendly Date format and dropped the 
DateFraction column.

[^2]:[Robert Shiller's Irrational Exuberance Data](
http://www.econ.yale.edu/~shiller/data.htm)

```{r echo=FALSE, Parse_Dates}
# Convert date to character format for parsing
cape$Date <- as.character(cape$Date)

# The as.Date function requires a day, so a .1 needs to be appended to the dates
# A 0 needs to be appended to October so that as.Date can differentiate it from
# January
october <- nchar(cape$Date) == 6
cape$Date <- ifelse(october,
                    paste(cape$Date, "0.1", sep = ""),
                    paste(cape$Date, ".1", sep = ""))

# Finally parse the result
cape$Date <- as.Date(cape$Date, '%Y.%m.%d')

# Rename the rows using the date column
#rownames(cape) <- cape$Date

# Drop the Date and DateFraction columns
cape <- subset(cape, select = -c(DateFraction))
```

The dataset contains the 10-Year Treasury Constant Maturity Rate labeled as
GS10. The rates are in percentage form, most likely to make them easier to
plot alongside the other features in the original excel file. I converted
them to decimal form to prepare for analysis.

```{r echo=FALSE, Fix_GS10}
cape$GS10 <- cape$GS10 / 100
```

## Feature Extraction

As an additional preparation step I created several new factors from the data:

1. Inflation
    + Measured as the annual % change in CPI
    + As a categorical value: Rising or Falling
2. Momentum
    + % Change in price over the past year
    + As a categorical value: Rising or Falling
3. Forward Returns
    + Measured as a % over 1, 3, 5 and 10 years
    + Categorical value representing if the 1 year "Outlook" was Bullish or 
    Bearish


Inflation was captured out of curiosity of its affect on future returns.
Momentum was measured because studies have shown that price advances and
declines can persist.[^3] I am  interested to see if it will make a good
complement to value in predicting returns.

I created a `pct_change` function to extract the above features from the 
price and CPI columns. I also created a function  to shift the results of the 
`pct_change` for building the future returns values.

Furthermore, I used categorical variations of the inflation and momentum to
assist in exploring positive and negative conditions of each.

[^3]:[Returns to Buying Winners and Selling Losers](
https://www.jstor.org/stable/2328882)

```{r echo=FALSE, pct_change_function}
pct_change <- function(vect, k = 12) {
  # Calculates the percent change over a specified lookback
  # :param vect:  vector of numbers
  # :param  k:    number of periods to lookback
  # :return:      running percent change over lookback length
  
  len = length(vect)
  
  # Create indexes that will be used to truncate the vector
  front_trunc <- (1:k)
  rear_trunc <- ((len-k+1):len)
  
  # Divide truncated versions of the vector by itself to get the percent change
  result <- vect[-front_trunc]/vect[-rear_trunc] - 1
  
  # Pad the results so that it matches the original vector length
  result <- c(rep(NA, k), result)
  
  return(result)
}
```

```{r echo=FALSE, Fwd_Return_Function}
fwd_return <- function(vect, k = 12) {
  # Shift a vector by k values
  # :param vect:  vector to shift
  # :param k:     number of periods to shift by
  # :return:      shifted vector

  len <- length(vect)

  p_changes <- pct_change(vect, k = k)

  # Remove the first k values of the vector
  result <- p_changes[-(1:k)]

  # Restore the vector to original length (NA is appended automatically)
  length(result) <- len

  return(result)
}

```


```{r echo=FALSE, Feature_Extraction}
# continuous measurments of growth and inflation
cape$Momentum <- pct_change(cape$Price)
cape$CPIChange <- pct_change(cape$CPI)

# nominal variation
growth <- cape$Momentum > 0
cape$Growth <- ifelse(growth, "Expansion", "Recession")
inflation <- cape$CPIChange > 0
cape$Inflation <- ifelse(inflation, "Inflation", "Deflation")
cape$Growth <- factor(cape$Growth)
cape$Inflation <- factor(cape$Inflation)

cape$Fwd1yrReturns <- fwd_return(cape$Price)
cape$Fwd3yrReturns <- fwd_return(cape$Price, k = 36)
cape$Fwd5yrReturns <- fwd_return(cape$Price, k = 60)
cape$Fwd10yrReturns <- fwd_return(cape$Price, k = 120)

bullish <- cape$Fwd1yrReturns > 0
cape$Outlook <- ifelse(bullish, "Bullish", "Bearish")
cape$Outlook <- factor(cape$Outlook)



```

# Univariate Plots Section

```{r echo=FALSE, Summary}
str(cape)
summary(cape)
```

The final dataset contains 1769 observations of 22 variables.

### Price

```{r echo=FALSE, message=FALSE, warning=FALSE, Price}
theme_set(theme_bw())
g <- ggplot(cape)

plt1 <- g + geom_line(aes(Date, Price))
plt2 <- g + geom_line(aes(Date, RealPrice))
plt3 <- plt1 + scale_y_log10() + labs(y = "Log Price")
plt4 <- plt2 + scale_y_log10() + labs(y = "Log RealPrice")

grid.arrange(plt1, plt2, plt3, plt4, ncol = 2)
```

Because the value of price is measured over time it is best plotted as a 
time series. The scale of the first plot makes it difficult to compare the data
starting in the late 1800s to more recent years. The dataset already contains a 
transformed version in the form of RealPrice, which has been adjusted for
inflation. Taking the log10 of the prices makes them even more readable, which
is a natural transformation because price growth is geometric. The final
inflation adjusted and log transformed graph makes it easy to compare the growth
and recessions over the entire history of the dataset.

### Dividends and Earnings

Logically, the dividends and earnings should benefit from the same
transformations.

```{r echo=FALSE, message=FALSE, warning=FALSE, Dividend}
plt1 <- g + geom_line(aes(Date, Dividend))
plt2 <- g + geom_line(aes(Date, RealDividend))
plt3 <- plt1 + scale_y_log10() + labs(y = "Log Dividend")
plt4 <- plt2 + scale_y_log10() + labs(y = "Log RealDividend")

grid.arrange(plt1, plt2, plt3, plt4, ncol = 2)
```

Dividends show the same steady growth over time as price. But the second half of 
the data appears to have much lower variance than the first half. I wonder if
this is due to the decision to start taxing dividends in 1954.[^4]

[^4]:[A brief history of dividend tax rates](
http://www.dividend.com/taxes/a-brief-history-of-dividend-tax-rates/)

```{r echo=FALSE, message=FALSE, warning=FALSE, Earnings}
plt1 <- g + geom_line(aes(Date, Earnings))
plt2 <- g + geom_line(aes(Date, RealEarnings))
plt3 <- plt1 + scale_y_log10() + labs(y = "Log Earnings")
plt4 <- plt2 + scale_y_log10() + labs(y = "Log RealEarnings")

grid.arrange(plt1, plt2, plt3, plt4, ncol = 2)
```

An interesting feature of the earnings data are the two significant dips that 
occurring during the Great Depression and Great Recession of the 1930's and 2008.
Earnings appear to have been impacted even harder than stock prices.

### Interest Rates

```{r echo=FALSE, message=FALSE, warning=FALSE, Interest_Rates}
plt1 <- g + geom_line(aes(Date, GS10)) + labs(title = "GS10 over time")
plt2 <- g + geom_density(aes(GS10)) + labs(title = "Distribution of GS10 rates")
plt3 <- plt2 + scale_x_log10() + labs(title = "", x = "Log10(GS10)")

#grid.arrange(plt1, plt2, plt3)
print(plt1)
print(plt2)
print(plt3)

summary(cape$GS10)
```

The 10-Year Treasury rates (GS10) show a huge spike in the 1970s as the Fed and
policy makers sought to curb runaway inflation. This creates a long right tail
in the distribution of rates. Transforming the scale to shows multiple peaks in
the distribution. The GS10 rate is heavily influenced by the Fed Funds rate.
Could these peaks be the default speeds of the Fed for boosting and taming the
economy?

### CAPE


```{r echo=FALSE, message=FALSE, warning=FALSE, CAPE}
plt1 <- g + geom_density(aes(CAPE))
plt2 <- plt1 + scale_x_log10(breaks = seq(0, 45, 5)) + labs(x = "LogCAPE")

grid.arrange(plt1, plt2)
```

CAPE, the Cyclically Adjusted Price to Earnings Ratio, is the heart of the 
dataset. Robert Shiller created the ratio to smooth for inflation and business
cycle affects by dividing the inflation adjusted price by 10 year average of 
the earnings.

Transforming to log scale makes it easier to see an important characteristic.
The tails of the distribution represent the cheapest and most expensive readings
for the market. However, the normal scale of the x axis diminishes the fact that
changes in the lower end represent proportionately larger percentage differences
in price per earnings. For example, the difference between a value of 4 and 5 
means paying 25% more for earnings, whereas 40 to 41 is only and increase of 
2.5%. The log scale balances the emphasis placed on the rare occurrence of 
valuations at both ends of the spectrum.

### Momentum

```{r echo=FALSE, message=FALSE, warning=FALSE, Momentum}
g + geom_density(aes(Momentum))
g + geom_boxplot(aes("", Momentum)) + coord_flip()
summary(cape$Momentum)
```

The Momentum distribution has long tails, with outlier years returning over 
100% and losing more than 50%. The boxplot makes these extreme occurrences easy
to see. The mean annual return was 
`r round(mean(cape$Momentum, na.rm = TRUE) * 100, 2) `%.

### Inflation

```{r echo=FALSE, message=FALSE, warning=FALSE, Inflation}
g + geom_density(aes(CPIChange))
summary(cape$CPIChange)
```

The distribution of CPIChange shows very high kurtosis, with a mean of 
`r round(mean(cape$CPIChange, na.rm = TRUE) * 100, 2)`%.
This aligns with the Fed's target, while the tails show just how far inflation
and deflation can go when they get out of hand.

```{r echo=FALSE, message=FALSE, warning=FALSE, Economic_Seasons}
# The first 12 rows contain NA values for both Inflation and growth
plt1 <- ggplot(cape[-(1:12),]) + geom_bar(aes(Inflation))

plt2 <- ggplot(cape[-(1:12),]) + geom_bar(aes(Growth))
grid.arrange(plt1, plt2, ncol = 2)
```

The bar charts show that the US has spent most of period in a state of rising
inflation and growth.

### Forward Returns

```{r echo=FALSE, message=FALSE, warning=FALSE, Forward_Returns}
plt1 <- g + geom_density(aes(Fwd1yrReturns))
plt2 <- g + geom_density(aes(Fwd3yrReturns))
plt3 <- g + geom_density(aes(Fwd5yrReturns))
plt4 <- g + geom_density(aes(Fwd10yrReturns))

grid.arrange(plt1, plt2, plt3, plt4, ncol = 2)
summary(cape[c("Fwd1yrReturns", "Fwd3yrReturns",
               "Fwd5yrReturns", "Fwd10yrReturns")])
```

>Note that the Fwd1yReturns are the same values as momentum, but shifted 1 year.

An interesting feature of the forward returns is that they become more
positively skewed the longer the duration. The skew is the affect of
compounding, but there is a lower limit due to prices having never been below 0.

A logical transformation is to convert all of the returns to their compound
annual growth rate (CAGR). This will also make it easier to compare them to
each other.

```{r echo=FALSE, message=FALSE, warning=FALSE, CAGR_Forward_Returns}
# this function will convert the returns to a CAGR
annualize <- function(r,t) {
  return((1 + r)^(1 / t) - 1)
}

plt1 <- g + geom_density(aes(Fwd1yrReturns))
plt2 <- g + geom_density(aes(annualize(Fwd3yrReturns, 3))) +
                                    labs(x = "3 Year CAGR")
plt3 <- g + geom_density(aes(annualize(Fwd5yrReturns, 5))) +
                                    labs(x = "5 Year CAGR")
plt4 <- g + geom_density(aes(annualize(Fwd10yrReturns, 10))) +
                                    labs(x = "10 Year CAGR")

grid.arrange(plt1, plt2, plt3, plt4, ncol = 2)
print("1 year forward returns")
summary(cape$Fwd1yrReturns)
print("3 year forward returns")
summary(annualize(cape$Fwd3yrReturns, 3))
print("5 year forward returns")
summary(annualize(cape$Fwd5yrReturns, 5))
print("10 year forward returns")
summary(annualize(cape$Fwd10yrReturns, 10))
```

The oft referred to fat tails of the returns are still present and a slight negative 
skew is visible. This is characteristic of the risks investors face; extreme 
returns occur more often than they would with a normal distribution, with 
greater than normal occurrence negative returns.

The mean annual returns decrease as the forward looking periods decrease. This
is a result of the negative skew and its affect on compounding. Large losses
require even larger gains for recovery, causing realized CAGRs to be lower than 
the mean Fwd1yrReturns.

After seeing the transformed values I decided to store them for easier access
during rest of the analysis.

```{r echo=FALSE, message=FALSE, warning=FALSE, Save_CAGR}
cape$Fwd3yrCAGR <- annualize(cape$Fwd3yrReturns, 3)
cape$Fwd5yrCAGR <- annualize(cape$Fwd5yrReturns, 5)
cape$Fwd10yrCAGR <- annualize(cape$Fwd10yrReturns, 10)

# g needs to be updated now that there are new features in the df
g <- ggplot(cape)
```

# Univariate Analysis

### What is the structure of your dataset?

The dataset contains 1769 months of data covering 22 features, one of which 
being the date. The data covers the price, dividends and earnings for the US
stock market since 1871. Also included are the Consumer Price Index (CPI) and
the 10-Year Treasury rates. The first allows conversion of values from nominal 
to inflation adjusted levels. The Treasury rates are commonly used as a risk
free benchmark.

There are also many NA values in the data because of the numerous values created
from running calculations. CAPE, for example, requires 10 years of earnings 
values to smooth before it begins showing values.

### What is/are the main feature(s) of interest in your dataset?

Shiller's CAPE ratio is the main feature of the dataset, allowing the analysis
of market valuations over the extended time period. I also introduced momentum
measures so that I can attempt to build a model for forecasting future returns.

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

CPI is crucial in transforming the data over time to account for inflation. In
addition, I expect to see that inflation itself will have affects on stock
returns. I believe this will also be the case for interest rates, which are 
often used by governments to boost economic growth.

### Did you create any new variables from existing variables in the dataset?

A measure of momentum and inflation were created to show their running 12 month
rates of change. Forward returns were also calculated and then transformed into
compound annual growth rates (CAGRs) during the analysis.

Categorical values were created to allow partitioning the data into periods of
Rising and Falling states of Growth and Inflation. An Outlook categorical 
variable was also created to test if predicting bullish and bearish periods was
more feasible than predicting the specific level of return.

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

Skewed distributions are the norm in financial data, starting with the long 
tails in returns and carried through to prices with the effect of compounding
over time. Interest rates and inflation also showed long tails.

Log transformations were applied to prices, dividends and earnings to reduce 
positive skew. They are also a natural transformation to apply because they 
convert the exponential growth of continuous compounding into an additive one. 
The CAPE ratio is generated from those variables and benefited from the same 
transformation.

I also took advantage of the inflation adjustments that were already in the data 
set to help visualize changes in the variables over time.

Last, I transformed the different forward return measures to annual measures to 
reduce skew and so that they could be directly compared.

# Bivariate Plots Section

### Stock Features

```{r echo=FALSE, message=FALSE, warning=FALSE, Stock_Features}
plt1 <- g + geom_point(aes(RealPrice, RealEarnings), alpha = 1/10)
plt2 <- g + geom_point(aes(RealPrice, RealDividend), alpha = 1/10)
plt3 <- g + geom_point(aes(RealDividend, RealEarnings), alpha = 1/10)

grid.arrange(plt1, plt2, plt3, ncol = 2)

```

```{r echo=FALSE, message=FALSE, warning=FALSE, Stock_Features_Alpha}
plt1 <- g + geom_point(aes(log10(RealPrice), 
                           log10(RealEarnings)), alpha = 1/10)
plt2 <- g + geom_point(aes(log10(RealPrice), 
                           log10(RealDividend)), alpha = 1/10)
plt3 <- g + geom_point(aes(log10(RealDividend),
                           log10(RealEarnings)), alpha = 1/10)

grid.arrange(plt1, plt2, plt3, ncol = 2)

```

No surprises here, all of the stock features are highly correlated. Applying
the log transform and transparency to reduce over plotting shows that they have 
strong linear relationships and have varied together over time.

### CAPE and Forward Returns
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_vs_returns}
# Convenience function created because several features will be plotted
# against forward returns
plot_vs_returns <- function(df, x, x_label, alpha = 1) {
  g <- ggplot(aes(x = x), data = df)
  plt1 <- g + geom_point(aes(y = Fwd1yrReturns), alpha = alpha) + labs(x = x_label)
  plt2 <- g + geom_point(aes(y = Fwd3yrCAGR), alpha = alpha) + labs(x = x_label)
  plt3 <- g + geom_point(aes(y = Fwd5yrCAGR), alpha = alpha) + labs(x = x_label)
  plt4 <- g + geom_point(aes(y = Fwd10yrCAGR), alpha = alpha) + labs(x = x_label)
  
  grid.arrange(plt1, plt2, plt3, plt4)
}
```


```{r echo=FALSE, message=FALSE, warning=FALSE, Cape_to_Returns}
plot_vs_returns(cape, cape$CAPE, "CAPE", alpha = 1/10)
```

There appears to be a negative correlation to forward returns, but the effect is
less prominent in the Fwd1yrReturns. It looks as though the CAPE might be less 
valuable in near term forecasting than for making long term predictions.

The plots also look like they could benefit from using the Log10 transformed
version of CAPE.

```{r echo=FALSE, message=FALSE, warning=FALSE, Normalized_Cape_to_Returns}
plot_vs_returns(cape, log10(cape$CAPE), "Log10(CAPE)", alpha = 1/10)
```

The `cor.test` function showed that he Pearson R correlations were -0.19, -0.30,
-0.37 and -0.40, in order of increasing forward looking periods.

The following plots break CAPE into quintiles and show the mean returns for 
each quintile.

```{r echo=FALSE, message=FALSE, warning=FALSE, Cape_to_Mean_Returns}
# This function is necessary to plot returns against cape, while also
# maintaining meaningful x lables for the cut groups
log10_cut <- function(x, breaks = 5, digits = 0) {
  # Same functionality as cut, but cuts along log adjusted values 
  # :param x:       values to cut
  # :param breaks:  number of cuts to make
  # :param digits:  number of decimal places to round the labels
  # :returns cut data, but with original values in lables
  cuts <- cut(log10(x), breaks = breaks)
  
  # sub function to apply
  level_mapper <- function(lev) {
    return(gsubfn("(\\d+.\\d+)",
                  ~ round(10 ^ as.numeric(x), digits = digits),
                  lev))
  }
  
  levels(cuts) <- sapply(levels(cuts), level_mapper)
  return(cuts)
}

g_cape <- ggplot(aes(x = log10_cut(CAPE)), 
                 data = subset(cape, !is.na(CAPE)))

g_cape + geom_bar(aes(y = Fwd1yrReturns), stat = "summary", fun.y = mean)
g_cape + geom_bar(aes(y = Fwd3yrCAGR), stat = "summary", fun.y = mean)
g_cape + geom_bar(aes(y = Fwd5yrCAGR), stat = "summary", fun.y = mean)
g_cape + geom_bar(aes(y = Fwd10yrCAGR), stat = "summary", fun.y = mean)
```

Average returns are higher starting from points when the market was selling at 
a discount. 1 yr returns show the highest average for the lowest CAPE reading. 
But the just looking at the mean does not say anything about the distribution of
those returns.

```{r echo=FALSE, message=FALSE, warning=FALSE, Cape_to_Returns_Boxplots}
g_cape + geom_boxplot(aes(y = Fwd1yrReturns))
g_cape + geom_boxplot(aes(y = Fwd3yrCAGR))
g_cape + geom_boxplot(aes(y = Fwd5yrCAGR))
g_cape + geom_boxplot(aes(y = Fwd10yrCAGR))
```

The boxplots show that the means are affected by skewed distributions. The
median returns are more of an indication of what investors would earn during
a typical year.

### Momentum and Forward Returns

```{r echo=FALSE, message=FALSE, warning=FALSE, Momentum_Returns}
plot_vs_returns(cape, cape$Momentum, "Momentum", alpha = 1/20)
```

There doesn't appear to be a strong correlation between Momentum and forward
returns. It looks like there are outliers. It might help to zoom in.

```{r echo=FALSE, message=FALSE, warning=FALSE, Momentum_ClearPlot}
plt1 <-ggplot(aes(Momentum, Fwd1yrReturns), data = cape) +
  geom_point(alpha = 1/20) + xlim(-.5, .7)
plt2 <-ggplot(aes(Momentum, Fwd3yrCAGR), data = cape) +
  geom_point(alpha = 1/20) + xlim(-.5, .7)
plt3 <-ggplot(aes(Momentum, Fwd5yrCAGR), data = cape) +
  geom_point(alpha = 1/20) + xlim(-.5, .7)
plt4 <-ggplot(aes(Momentum, Fwd10yrCAGR), data = cape) +
  geom_point(alpha = 1/20) + xlim(-.5, .7)

grid.arrange(plt1, plt2, plt3, plt4)
```

There still doesn't appear to be a correlation, which surprises me. Momentum is
the tool of choice of trend followers. Maybe the categorical Growth variable 
will shed some light.

```{r echo=FALSE, message=FALSE, warning=FALSE, Growth_to_CAGR}
# trim the first 12 rows to remove NA values
plt1 <- ggplot(aes(Growth, Fwd1yrReturns), data = cape[-(1:12),]) + geom_boxplot()
plt2 <- ggplot(aes(Growth, Fwd3yrCAGR), data = cape[-(1:12),]) + geom_boxplot()
plt3 <- ggplot(aes(Growth, Fwd5yrCAGR), data = cape[-(1:12),]) + geom_boxplot()
plt4 <- ggplot(aes(Growth, Fwd10yrCAGR), data = cape[-(1:12),]) + geom_boxplot()

grid.arrange(plt1, plt2, plt3, plt4)
```

Separating the data into Rising and Falling Growth periods does not appear to
provide any benefit at all in predicting higher returns. However, it does
tighten the distribution and reduce the tails of the forward 1 year returns, a
particular benefit in reducing risk. This is confirmation of something I have
learned from studying trend following  in the past: its main benefit is not in
finding higher returns, but in avoiding the worst periods. However, I was
expecting at least some correlation to higher future returns. If momentum were
the focus of this study I would consider looking at shorter forward looking
periods.

One thing to note is that this connection reverses after the first year, with
the 3 and 5 year plots showing the opposite connection and lower average returns
after Rising Growth readings.

### Interest Rates

```{r echo=FALSE, message=FALSE, warning=FALSE, GS10_to_returns}
plot_vs_returns(cape, cape$GS10, "GS10", alpha = 1/10)
```

It looks like there could be correlation over longer time periods. For 1 year 
the Pearson R is only .05, but it increases to .14, .21 and .39 over the 
respective longer periods.

```{r echo=FALSE, message=FALSE, warning=FALSE, GS10_CAPE}
ggplot(cape[-(1:12),]) + geom_point(aes(GS10, log10(CAPE)), alpha = 1/10)
cor.test(cape$GS10, cape$CAPE)
cor.test(cape$GS10, log10(cape$CAPE))
```

It does look like there could be some negative correlation between GS10 rates 
and CAPE. But the `cor.test` function returned -.16 indicating that it
is very weak. The correlation is slightly stronger at -.22 when using the log10
transformed values of CAPE.

### Inflation

```{r echo=FALSE, message=FALSE, warning=FALSE, Inflation_to_returns}
plot_vs_returns(cape, cape$CPIChange, "Inflation", alpha = 1/10)
```

Although it is not readily apparent in the plot, the `cor.test` function found 
a weak correlation of .36 between CPIChange and Fwd10yrCAGR. It looks like 
inflation only correlates with returns over long periods.

```{r echo=FALSE, message=FALSE, warning=FALSE, te}
g + geom_point(aes(CPIChange, Momentum), alpha = 1/10)
```

Inflation does not show correlation to Momentum either.

```{r echo=FALSE, message=FALSE, warning=FALSE, Inflation_to_GS10}
g + geom_point(aes(CPIChange, GS10), alpha = 1/10)
```

No correlation was visible between Inflation( as CPIChange) and the GS10 rates. 
I was expecting to find one because of the Federal Reserve's use of rate policy 
to manage the inflation rate. It's possible that there is a delay between market
changes and policy reactions, and that lagging one against the other would 
reveal correlation.

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

Forward returns were negatively correlated with CAPE. They did not show 
correlation to Momentum, Inflation or the GS10 rates. However, their 
distribution did appear tighter in the 1 year forward returns following rising
momentum.

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

Price, Dividend and Earnings were all highly correlated. This was expected as
it was visible in the time series plots that they have varied together over the
duration of the dataset.

CAPE and GS10 had a weak negative correlation. This could be the affect of 
investors being willing to pay a higher price for returns during low interest 
rate periods, and vice versa.

### What was the strongest relationship you found?

The strongest relationship was between CAPE and forward returns, with the 
correlation increasing over the longer periods tested.

# Multivariate Plots Section

### Growth and CAPE vs Foward Returns

```{r echo=FALSE, message=FALSE, warning=FALSE, Growth_CAPE}
ggplot(cape[-(1:12),]) + geom_point(aes(CAPE, Fwd1yrReturns, col = Growth)) +
  scale_x_log10()
```

It's difficult to see in this plot, but it does look like there is a stronger
trend between CAPE and forward 1 year returns during periods of recession.

```{r echo=FALSE, message=FALSE, warning=FALSE, Growth_CAPE_Facet_Wrap}
ggplot(cape[-(1:12),]) + geom_point(aes(CAPE, Fwd1yrReturns, col = Growth)) +
  scale_x_log10() + facet_wrap(~Growth)
```

Using `facet_wrap` to separate the conditions makes the difference easier to 
see. Periods of recession in the stock market exhibit a stronger negative
correlation between CAPE and price than during expansion.

```{r echo=FALSE, message=FALSE, warning=FALSE, CAGR_Growth_CAPE_Facet_Wrap}
ggplot(cape[-(1:12),]) + geom_point(aes(CAPE, Fwd3yrCAGR, col = Growth)) +
  scale_x_log10() + facet_wrap(~Growth)
ggplot(cape[-(1:12),]) + geom_point(aes(CAPE, Fwd5yrCAGR, col = Growth)) +
  scale_x_log10() + facet_wrap(~Growth)
ggplot(cape[-(1:12),]) + geom_point(aes(CAPE, Fwd10yrCAGR, col = Growth)) +
  scale_x_log10() + facet_wrap(~Growth)
```

The feature combination doesn't appear to have the same benefit over longer time
periods.

### CAPE and Inflation to Returns

```{r echo=FALSE, message=FALSE, warning=FALSE, CAPE_Inflation_Returns}
ggplot(cape[-(1:12),]) + geom_point(aes(CAPE, Fwd1yrReturns, col = Inflation)) +
  scale_x_log10() + facet_wrap(~Inflation)
ggplot(cape[-(1:12),]) + geom_point(aes(CAPE, Fwd3yrCAGR, col = Inflation)) +
  scale_x_log10() + facet_wrap(~Inflation)
ggplot(cape[-(1:12),]) + geom_point(aes(CAPE, Fwd5yrCAGR, col = Inflation)) +
  scale_x_log10() + facet_wrap(~Inflation)
ggplot(cape[-(1:12),]) + geom_point(aes(CAPE, Fwd10yrCAGR, col = Inflation)) +
  scale_x_log10() + facet_wrap(~Inflation)
```

Inflation and Deflation also appear to complement CAPE's correlation to returns.

The numeric measures of Inflation and Growth did not show correlation during 
bivariate analysis, suggesting that they should work well together to enhance
the power of CAPE.

```{r echo=FALSE, message=FALSE, warning=FALSE, Quad_Wrap}
g_cape + geom_boxplot(aes(y = Fwd1yrReturns)) + 
  facet_wrap(Growth~Inflation)
```

CAPE looks strongest in predicting forward 1 year returns during deflationary
recession years.

### The search for yield

The inverse of CAPE can be interpreted as a yield similar to interest rates. Are 
investors willing to pay more for earnings when interest rates are low?

```{r echo=FALSE, message=FALSE, warning=FALSE, Dividend_Yield}
g + geom_density(aes(1/CAPE - GS10)) +
  labs(title = "Stock Yield: Inverse CAPE minus the risk free rate")
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Modified_CAPE_vs_Returns}
plot_vs_returns(cape, 1/cape$CAPE - cape$GS10,
                "Inverse CAPE - GS10", alpha = 1/10)
```

There does appear to be a positive correlation between the excess yield of 
stocks and forward returns.

```{r echo=FALSE, message=FALSE, warning=FALSE, LogModified_CAPE_vs_Returns}
plot_vs_returns(cape, log10(1/cape$CAPE) - cape$GS10,
                "Log10(Inverse CAPE) - GS10", alpha = 1/10)
```

Using the log10 inverse of CAPE makes the relationship look even stronger. The
Pearson R correlations between the new value and forward returns
are .18, .28, .34 and .35, indicating that this combination does not provide any
benefit. These values are slightly lower than using CAPE alone.

### Prediciting the future

I will use `lm` to fit a linear model to predict future returns. 

```{r echo=FALSE, message=FALSE, warning=FALSE, linear_models}
lm1 <- lm(Fwd1yrReturns ~ log10(CAPE) + Growth + Inflation + GS10,
          data = cape)
lm3 <- lm(Fwd3yrCAGR ~ log10(CAPE) + Growth + Inflation + GS10,
          data = cape)
lm5 <- lm(Fwd5yrCAGR ~ log10(CAPE) + Growth + Inflation + GS10,
          data = cape)
lm10 <- lm(Fwd10yrCAGR ~ log10(CAPE) + Growth + Inflation + GS10,
          data = cape)

tm1 <- rpart(Outlook ~ Momentum, data = cape)

mtable(
  "Model 1" = lm1,
  "Model 2" = lm3,
  "Model 3" = lm5,
  "Model 4" = lm10
)
```

The strongest predictions were for 10 year forward returns, but with only 35%
of the variance in returns being explained by the model.

A classification model can also be used to predict the Outlook variable. I used
`rpart` to create a tree-based model to predict if the next year would be
bullish or bearish.

```{r echo=FALSE, message=FALSE, warning=FALSE, Classification_model}
# build the classification model
tm <- rpart(Outlook ~ log10(CAPE) + Growth + Inflation + GS10,
            data = cape)

# capture predictions
pred <- predict(tm, cape, type = "class")

# calculate accuracy using the confusion matrix

confmat <- table(cape$Outlook, pred)
print(confmat)

accuracy <- sum(diag(confmat)/sum(confmat))

```

The model was 78% accurate in labeling the dataset. One way of measuring the
power of a classification model is to compare it to a majority classifier, 
which would have labelled all of the data as bullish with 63% accuracy.

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

The Growth and Inflation features both strengthened the correlation between
CAPE and 1 year forward returns. CAPE appears to have the strongest
correlation to forward returns after periods of recession and deflation.
One possible explanation is that periods of expanding growth and inflation
push all prices up, without discrimination to the fundamental value of the
market. In the reverse situation, a benefit of this relationship is that
CAPE can be used to help identify good opportunities after stock and 
consumer prices have been falling.

### Were there any interesting or surprising interactions between features?

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

Linear models were used to predict future returns. The models showed that 
returns are easier to predict over longer time horizons, but even the 10 year
model was only able to predict 35% of the variation in returns.

A classification model was able to predict the 1 year outlook with 78% accuracy.
This showed some strength over the majority classier.

To settle curiosity, the models are 71% confident that the next year will be 
bullish, but are only expecting a 1% return over the next 10 years (as of
May 2018).

One caveat to keep in mind is that I did not split the data into training and
test sets to control for over-fitting. These models were purely for exploring
relationships in the data.

------

# Final Plots and Summary


### Plot One
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_One}
theme_set(theme_classic())

g + geom_line(aes(Date, RealPrice)) + ggtitle("Stock Market Prices") +
  scale_y_log10()
```

### Description One

This plot shows the log adjusted change in stock prices since 1871. Stocks have
been a good investment, but the ride has been anything but smooth.

### Plot Two
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Two}

ggplot(aes(x = log10_cut(CAPE), y = Fwd10yrCAGR),
                    data = subset(cape, !is.na(CAPE))) +
  labs(x = "CAPE") + 
  scale_y_continuous(labels = scales::percent, limits = c(-.1, .2)) +
  geom_boxplot(aes(fill = log10_cut(CAPE))) +
  labs(title = "Future Returns", fill = "CAPE")
```

### Description Two

Market valuations have power in predicting future returns. The average 10 year
returns for the stock market are higher and with less variance starting from
low CAPE readings. Coincidentally, the CAPE ratio is currently at 
`r round(cape$CAPE[nrow(cape)])` as of `r cape$Date[nrow(cape)]`. Current 
valuations suggest that the 10 year outlook for stocks is grim, which runs
contrary to public opinion.


### Plot Three
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Three}
ggplot(subset(cape, !is.na(Growth))) +
  geom_boxplot(aes(Growth, Fwd1yrReturns, fill = Growth), alpha = 0.5) + 
  ggtitle("Trimming the fat tails of returns")
```

### Description Three

Periods of expansion have less variance in returns than those of recession. This
is a subtle, but important effect. Most investors have a hard time dealing with
volatility. They can benefit from an indicator letting them know when to step 
aside or increase diversification (i.e. tilt portfolio allocations to bonds).

------

# Reflection

This project gave me the opportunity to explore an interesting dataset 
covering over 100 years of stock market data. Some wrangling was required to
prepare the data and create additional features of interest. I began by exploring
all of the variables, creating a few interesting ones along the say. Then I 
analyzed the interactions between the features and finally created a few models
for predicting stock market action.

Several variables were found to be correlated to future returns. These included
the CAPE valuation ratio, interests rates and inflation. However, none of the 
correlations were very strong. This made building an accurate model for 
predicting the future returns difficult, with the strongest model only 
predicting 35% of the variance of returns. On the other hand, predicting 
whether the next year was bullish or bearish proved easier. A classification
tree model was able to predict this feature with 78% accuracy.

This analysis could be enhanced by introducing additional asset classes to the
dataset and building a model that created actionable predictions. For example,
a model could be created to predict which asset would have the best performance
and used to guide portfolio allocation.

